---
title: "R-Analysis Reverse"
author: "Dina Sami -  Heba Amer"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    
    toc: true
    highlight: "default"
    number_sections: true
    theme: "readable"
    toc_depth: 3
    toc_float: true
---


# Simplifying the Table


```{r read-data }

gene_counts <- read.table("gene_counts.txt", header = TRUE, sep = "\t", row.names = 1)

colnames(gene_counts)
```


```{r remove-columns}

gene_counts <- gene_counts[, -c(1:5)]

colnames(gene_counts)
```

```{r rename-columns}

new_colnames <- sub(".*_(S[0-9]+)_.*", "\\1", colnames(gene_counts))

colnames(gene_counts) <- new_colnames

colnames(gene_counts)
```


```{r split-tables}
# Create a mapping of sample IDs to conditions with "Control"
sample_conditions <- data.frame(
  Sample = c("S25", "S26", "S27", "S28", "S29", "S30", "S31", "S33", "S34", "S35", "S36", "S37", "S38", "S39", "S40", "S41", "S42", "S43", "S44", "S45", "S46", "S47", "S48", "S49", "S50"),
  Condition = c("EtOH", "Control", "Control", "Control", "NASH", "NASH", "Control", "NASH", "NASH", "EtOH", "EtOH", "NASH", "NASH", "EtOH", "EtOH", "EtOH", "EtOH", "Control", "NASH", "NASH", "NASH", "EtOH", "Control", "EtOH", "EtOH")
)

# Subset the gene_counts data frame based on conditions
EtOH_samples <- sample_conditions$Sample[sample_conditions$Condition == "EtOH"]
Control_samples <- sample_conditions$Sample[sample_conditions$Condition == "Control"]
NASH_samples <- sample_conditions$Sample[sample_conditions$Condition == "NASH"]

gene_counts_EtOH <- gene_counts[, EtOH_samples]
gene_counts_Control <- gene_counts[, Control_samples]
gene_counts_NASH <- gene_counts[, NASH_samples]

# Display the first few rows of each subset
list(EtOH = head(gene_counts_EtOH), Control = head(gene_counts_Control), NASH = head(gene_counts_NASH))
```

```{r rename-columns-with-prefixes}
# Rename columns for each condition
colnames(gene_counts_Control) <- paste0("C_", colnames(gene_counts_Control))
colnames(gene_counts_EtOH) <- paste0("E_", colnames(gene_counts_EtOH))
colnames(gene_counts_NASH) <- paste0("N_", colnames(gene_counts_NASH))

# Display the column names of each subset
list(Control = colnames(gene_counts_Control), EtOH = colnames(gene_counts_EtOH), NASH = colnames(gene_counts_NASH))
```


```{r merge-and-save-all}
# Merge Control, EtOH, and NASH data frames
gene_counts_all <- cbind(gene_counts_Control, gene_counts_EtOH, gene_counts_NASH)

# Display the first few rows of the merged data frame
head(gene_counts_all)

# Save the merged data frame to a file
write.table(gene_counts_all, "raw_gene_counts_all.txt", sep = "\t", quote = FALSE, col.names = NA)
```

```{r}
# Define the columns you want to keep
columns_to_keep <- c("C_S27", "C_S28", "C_S31", "C_S43", "C_S48", 
                     "E_S36", "E_S41", "E_S42", "E_S49", "E_S50", 
                     "N_S29", "N_S30", "N_S37", "N_S38", "N_S46")

# Subset the dataframe to keep only the specified columns
gene_counts_subset <- gene_counts_all[, columns_to_keep]


# Save the merged data frame to a file
write.table(gene_counts_subset, "gene_counts_all.txt", sep = "\t", quote = FALSE, col.names = NA)


```




# Introduction

This analysis focuses on gene expression differences between Control, EtOH, and NASH samples using DESeq2. We will load the gene count data, perform filtering, and run the DESeq2 analysis step by step.

# Gene Counts Analysis

## Library Installation and Loading

The following block installs necessary libraries (if not already installed) and loads them for subsequent use. It also includes the installation of annotation databases for gene mapping.

```{r setup-installation, results="hide", warning=FALSE, message=FALSE}
# Install necessary libraries if not already installed
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(c("DESeq2", "pheatmap", "ggplot2", "EnhancedVolcano","AnnotationDbi","org.Hs.eg.db","biomaRt", "clusterProfiler"))

# Load necessary libraries
library(DESeq2)
library(pheatmap)
library(ggplot2)
library(EnhancedVolcano)
library("org.Hs.eg.db")
library(clusterProfiler)

```

## Load the Gene Count Data

In this step, we will load the gene count data from a tab-separated file called `gene_counts_all.tsv`. The data frame contains gene counts for each sample, with genes as rows and samples as columns.


```{r load-data, warning=FALSE, message=FALSE}

# Load the gene counts data
gene_counts <- read.table("gene_counts_all.txt", header = TRUE, sep = "\t", row.names = 1)

# Display the first few rows of the data frame
colnames(gene_counts)
```

## Prepare Metadata

This step involves creating a metadata (colData) data frame that describes the condition for each sample. The conditions include Control, EtOH, and NASH samples.

```{r prepare-metadata, warning=FALSE, message=FALSE}
# Create a metadata (colData) data frame that describes the conditions for each sample
sample_conditions <- data.frame(
  row.names = colnames(gene_counts),
  condition = c(rep("Control", 5), rep("EtOH", 5), rep("NASH", 5))
)

# Display the metadata
sample_conditions
```


## Filter Out Zero Counts

Here, we filter out genes with all zero counts across all samples, as they provide no information for analysis.

```{r filter-zero-counts, warning=FALSE, message=FALSE}
# Filter out genes with all zero counts
gene_counts_filtered <- gene_counts[rowSums(gene_counts != 0) > 0, ]

# Display the first few rows of the filtered data frame
dim(gene_counts_filtered)
```



## Filter by Mean Expression

After removing genes with zero counts, we further filter genes based on their mean expression across samples. Only genes with a mean expression greater than a specified threshold 10 are kept for analysis.

```{r filter-by-mean, warning=FALSE, message=FALSE}
# Calculate the mean expression for each gene
gene_means <- rowMeans(gene_counts_filtered)

# Filter out genes with mean expression below a certain threshold (e.g., mean < 10)
mean_threshold <- 10
gene_counts_filtered <- gene_counts_filtered[gene_means >= mean_threshold, ]

# Display the first few rows of the filtered data frame
dim(gene_counts_filtered)
```


## Create DESeq2 Dataset

DESeq2 requires a DESeqDataSet object, which is created by providing the count data and sample information (metadata). Here, we create the dataset using the filtered gene count data.

```{r create-deseq-dataset-filtered, warning=FALSE, message=FALSE}

# Create the DESeqDataSet with counts
dds_filtered <- DESeqDataSetFromMatrix(countData = gene_counts_filtered,
                                       colData = sample_conditions,
                                       design = ~ condition)


# Display the DESeqDataSet object
dds_filtered

```

## Run DESeq2

Now that we have the dataset ready, we run the DESeq2 analysis to identify differentially expressed genes between the conditions.

```{r run-deseq2, warning=FALSE, message=FALSE}
# Run the DESeq2 differential expression analysis
dds_filtered <- DESeq(dds_filtered)

dds_filtered
```


## Extract Results for Control vs EtOH

We are extracting the results of the differential expression analysis for the Control vs EtOH comparison.

```{r extract-results-etoh, warning=FALSE, message=FALSE}

# Extract the results for EtOH vs Control
results_etoh <- results(dds_filtered, contrast = c("condition", "Control", "EtOH"))

# Display the first few rows of the results
summary(results_etoh)


```

## Extract Results for NASH vs Control

We are extracting the results of the differential expression analysis for the NASH vs Control comparison.

```{r extract-results-nash, warning=FALSE, message=FALSE}

#Extract results for NASH:

results_nash <- results(dds_filtered, contrast = c("condition", "Control", "NASH"))


summary(results_nash)


```

# Quality Check for Data and Analysis

## Plot Dispersion Estimates

We are plotting the dispersion estimates to visualize gene-wise dispersion across samples.

```{r plot-dispersion, warning=FALSE, message=FALSE}

#Plot the dispersion estimates:

plotDispEsts(dds_filtered)

pdf("DispersionPlot.pdf")
plotDispEsts(dds_filtered)
dev.off()
```

## PCA Plot


```{r,warning=FALSE, message=FALSE}
# Perform variance-stabilizing transformation (VST) on dds_filtered
vsd <- vst(dds_filtered, blind = FALSE)  # or use rlog for rlog-transformed data

# Extract the transformed expression data
pca_data <- assay(vsd)  # This will be used for PCA


# Perform PCA on the transposed matrix (samples as rows)
pca <- prcomp(t(pca_data), scale. = TRUE)

# Extract the sample metadata (conditions)
sample_conditions <- colData(dds_filtered)$condition  # Assuming this contains 'Control', 'EtOH', 'NASH'

# Create a data frame for PCA plotting
pca_df <- data.frame(PC1 = pca$x[,1], PC2 = pca$x[,2], Condition = sample_conditions)

# Load ggplot2 for plotting
library(ggplot2)

# Plot PCA with points colored by Condition (Control, EtOH, NASH)
pca_plot <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Condition)) +
    geom_point(size = 3) +
    theme_minimal() +
    labs(title = "PCA of Gene Expression Data",
         x = paste0("PC1: ", round(100 * summary(pca)$importance[2,1], 1), "% variance"),
         y = paste0("PC2: ", round(100 * summary(pca)$importance[2,2], 1), "% variance")) +
    theme(plot.title = element_text(hjust = 0.5))

# Display the plot
print(pca_plot)

# Save the plot as a PDF
ggsave("PCA_plot_Control_EtOH_NASH.pdf", plot = pca_plot, width = 8, height = 6)
```

# Obtaining the Results

## Extract Significant Results for EtOH

We are extracting the significantly differentially expressed genes for EtOH vs Control (adjusted p-value \< 0.05).

```{r extract-significant-etoh, warning=FALSE, message=FALSE}

#Extract significant results:

sigs_etoh <- na.omit(results_etoh)
sigs_etoh <- sigs_etoh[sigs_etoh$padj < 0.05, ]


# Convert sigs_control_etoh (DESeqResults object) to a data frame
sigs_etoh_df <- as.data.frame(sigs_etoh)

# Load biomaRt package
library(biomaRt)

# Use the ENSEMBL database to retrieve gene symbols

# Try different Ensembl mirrors
mirrors <- c("www", "uswest", "useast", "asia")
mart <- NULL  # Initialize the mart variable

# Attempt to connect to each mirror
for (mirror in mirrors) {
    cat("Trying mirror:", mirror, "\n")
    try({
        mart <- useEnsembl(biomart = "ensembl", dataset = "hsapiens_gene_ensembl", mirror = mirror)
        if (!is.null(mart)) {
            cat("Connected to mirror:", mirror, "\n")
            break
        }
    }, silent = TRUE)
}

# If none of the mirrors work, check for updates
if (is.null(mart)) {
    cat("Unable to connect to any Ensembl mirror. Please check your internet connection or try updating Bioconductor packages.\n")
    # Optionally, you can add code to update Bioconductor packages here
}



# Get the gene symbols for the ENSEMBL IDs in sigs_control_etoh_df
genes <- rownames(sigs_etoh_df)
gene_info <- getBM(filters = "ensembl_gene_id", 
                   attributes = c("ensembl_gene_id", "hgnc_symbol"), 
                   values = genes, 
                   mart = mart)

# Add the ensembl_gene_id as a new column in sigs_control_nash_df
sigs_etoh_df$ensembl_gene_id <- rownames(sigs_etoh_df)

# Merge the gene symbols into sigs_control_nash_df
sigs_etoh_annotated <- merge(sigs_etoh_df, gene_info, by = "ensembl_gene_id", all.x = TRUE)

# Reorder columns to make gene symbols appear first if desired
sigs_etoh_annotated <- sigs_etoh_annotated[, c("hgnc_symbol", "ensembl_gene_id", colnames(sigs_etoh_annotated)[-c(1, ncol(sigs_etoh_annotated))])]


# Save the annotated results as an Excel file
library(writexl)

write_xlsx(sigs_etoh_annotated, "r_sigs_etoh_with_gene_symbols.xlsx")

```




```{r, warning=FALSE, message=FALSE}
# Install readxl if you haven't already
if (!requireNamespace("readxl", quietly = TRUE)) {
  install.packages("readxl")
}

library(readxl)

# Read your Excel file
sigs_e <- "r_sigs_etoh_with_gene_symbols.xlsx"
Sigs_e_1 <- read_excel(sigs_e)

# Create a preview
head(Sigs_e_1, 10)



```





## Extract Significant Results for NASH

We are extracting the significantly differentially expressed genes for NASH vs Control (adjusted p-value \< 0.05).

```{r extract-significant-nash, warning=FALSE, message=FALSE}

#Extract significant results:

sigs_nash <- na.omit(results_nash)
sigs_nash <- sigs_nash[sigs_nash$padj < 0.05, ]


# Convert sigs_control_nash (DESeqResults object) to a data frame
sigs_nash_df <- as.data.frame(sigs_nash)

# Load biomaRt package
library(biomaRt)

# Use the ENSEMBL database to retrieve gene symbols

# Try different Ensembl mirrors
mirrors <- c("www", "uswest", "useast", "asia")
mart <- NULL  # Initialize the mart variable

# Attempt to connect to each mirror
for (mirror in mirrors) {
    cat("Trying mirror:", mirror, "\n")
    try({
        mart <- useEnsembl(biomart = "ensembl", dataset = "hsapiens_gene_ensembl", mirror = mirror)
        if (!is.null(mart)) {
            cat("Connected to mirror:", mirror, "\n")
            break
        }
    }, silent = TRUE)
}

# If none of the mirrors work, check for updates
if (is.null(mart)) {
    cat("Unable to connect to any Ensembl mirror. Please check your internet connection or try updating Bioconductor packages.\n")
    # Optionally, you can add code to update Bioconductor packages here
}


# Get the gene symbols for the ENSEMBL IDs in sigs_control_nash_df
genes <- rownames(sigs_nash_df)
gene_info <- getBM(filters = "ensembl_gene_id", 
                   attributes = c("ensembl_gene_id", "hgnc_symbol"), 
                   values = genes, 
                   mart = mart)

# Add the ensembl_gene_id as a new column in sigs_control_nash_df
sigs_nash_df$ensembl_gene_id <- rownames(sigs_nash_df)

# Merge the gene symbols into sigs_control_nash_df
sigs_nash_annotated <- merge(sigs_nash_df, gene_info, by = "ensembl_gene_id", all.x = TRUE)

# Reorder columns to make gene symbols appear first if desired
sigs_nash_annotated <- sigs_nash_annotated[, c("hgnc_symbol", "ensembl_gene_id", colnames(sigs_nash_annotated)[-c(1, ncol(sigs_nash_annotated))])]


# Save the annotated results as an Excel file
library(writexl)
write_xlsx(sigs_nash_annotated, "r_sigs_nash_with_gene_symbols.xlsx")

```



```{r,warning=FALSE, message=FALSE }
# Install readxl if you haven't already
if (!requireNamespace("readxl", quietly = TRUE)) {
  install.packages("readxl")
}

library(readxl)

# Read your Excel file
sigs_n <- "r_sigs_nash_with_gene_symbols.xlsx"
Sigs_n_1 <- read_excel(sigs_n)

# Create a preview
head(Sigs_n_1, 10)


```





# Visualise the Results



## Final Volcano Plot for EtOH with Gene Symbols

We are generating a Volcano plot using gene symbols for labeling the significant results in EtOH vs Control.

```{r final-volcano-etoh, warning=FALSE, message=FALSE}

#Map ENSEMBL IDs to gene symbols and generate the Volcano plot:

sigs.df_etoh <- as.data.frame(sigs_etoh)

sigs.df_etoh$symbol <- mapIds(org.Hs.eg.db, 
                         keys = rownames(sigs.df_etoh), 
                         keytype = "ENSEMBL", 
                         column = "SYMBOL",
                         multiVals = function(x) paste(x, collapse = ";"))


# Define the cutoffs
pCutoffValue <- 0.05
FCcutoffValue <- 1.0


EnhancedVolcano(sigs.df_etoh,
                lab = sigs.df_etoh$symbol,
                x = 'log2FoldChange',
                y = 'padj',
                pCutoff = pCutoffValue,
                FCcutoff = FCcutoffValue,
                pointSize = 2.0,
                labSize = 3.0,
                title = 'Volcano Plot with Gene Symbols',
                subtitle = 'EtOH Significant Results',
                caption = paste('pCutoff:', pCutoffValue, 'FCcutoff:', FCcutoffValue))

# Save the EnhancedVolcano plot as a PDF
pdf("R_EtOH_EnhancedVolcanoPlot_with_GeneSymbols.pdf")
EnhancedVolcano(sigs.df_etoh,
                lab = sigs.df_etoh$symbol,  # Use the gene symbols for labeling
                x = 'log2FoldChange',
                y = 'padj',
                pCutoff = pCutoffValue,
                FCcutoff = FCcutoffValue,
                pointSize = 2.0,
                labSize = 3.0,
                title = 'Volcano Plot with Gene Symbols',
                subtitle = 'EtOH Significant Results',
                caption = paste('pCutoff:', pCutoffValue, 'FCcutoff:', FCcutoffValue))
dev.off()

```

## Final Volcano Plot for NASH with Gene Symbols

We are generating a Volcano plot using gene symbols for labeling the significant results in NASH vs Control.

```{r final-volcano-nash, warning=FALSE, message=FALSE}

# Map ENSEMBL IDs to gene symbols and generate the Volcano plot:

sigs.df_nash <- as.data.frame(sigs_nash)

sigs.df_nash$symbol <- mapIds(org.Hs.eg.db, 
                         keys = rownames(sigs.df_nash), 
                         keytype = "ENSEMBL", 
                         column = "SYMBOL",
                         multiVals = function(x) paste(x, collapse = ";"))


# Define the cutoffs
pCutoffValue <- 0.05
FCcutoffValue <- 1.0

EnhancedVolcano(sigs.df_nash,
                lab = sigs.df_nash$symbol,
                x = 'log2FoldChange',
                y = 'padj',
                pCutoff = pCutoffValue,
                FCcutoff = FCcutoffValue,
                pointSize = 2.0,
                labSize = 3.0,
                title = 'Volcano Plot with Gene Symbols',
                subtitle = 'NASH Significant Results',
                caption = paste('pCutoff:', pCutoffValue, 'FCcutoff:', FCcutoffValue))

# Save the EnhancedVolcano plot as a PDF
pdf("R_NASH_EnhancedVolcanoPlot_with_GeneSymbols.pdf")
EnhancedVolcano(sigs.df_nash,
                lab = sigs.df_nash$symbol,  # Use the gene symbols for labeling
                x = 'log2FoldChange',
                y = 'padj',
                pCutoff = pCutoffValue,
                FCcutoff = FCcutoffValue,
                pointSize = 2.0,
                labSize = 3.0,
                title = 'Volcano Plot with Gene Symbols',
                subtitle = 'NASH Significant Results',
                caption = paste('pCutoff:', pCutoffValue, 'FCcutoff:', FCcutoffValue))
dev.off()

```


## Top Genes Heatmap Control_EtOH


```{r,warning=FALSE, message=FALSE}

# Assuming you already have `sigs_control_etoh` with gene symbols loaded
# Load the DESeq2 object that contains the count data (dds_filtered object)
# Filter the DESeq2 object to only keep Control and EtOH samples
dds_filtered_control_etoh <- dds_filtered[, colData(dds_filtered)$condition %in% c("Control", "EtOH")]

# Get the normalized counts for the significant genes from Control vs. EtOH
# Extract only the genes that are significant and present in your filtered data
significant_genes_e <- rownames(sigs_etoh)  # Assuming rownames are gene symbols or ENSEMBL IDs
normalized_counts <- counts(dds_filtered_control_etoh, normalized = TRUE)
sig_gene_counts_e <- normalized_counts[significant_genes_e, ]


# Step 1: Sort the significant genes from EtOH by log2FoldChange and select the top 20

# Select top 10 upregulated genes (highest log2FoldChange)
top_up_e <- sigs_etoh_df[order(-sigs_etoh_df$log2FoldChange), ][1:10, ]

# Select top 10 downregulated genes (lowest log2FoldChange)
top_down_e <- sigs_etoh_df[order(sigs_etoh_df$log2FoldChange), ][1:10, ]

# Combine top up- and downregulated genes
top_genes_e <- rbind(top_up_e, top_down_e) 

# Step 2: Get the normalized counts for these top genes
top_gene_ensembl_ids_e <- rownames(top_genes_e)  # These are ENSEMBL IDs
top_gene_counts_e <- normalized_counts[top_gene_ensembl_ids_e, ]

# Step 3: Use biomaRt to convert ENSEMBL IDs to gene symbols for EtOH


# Query to get gene symbols for the given ENSEMBL IDs
gene_mapping_e <- getBM(
    filters = "ensembl_gene_id",
    attributes = c("ensembl_gene_id", "hgnc_symbol"),
    values = top_gene_ensembl_ids_e,
    mart = mart
)

# Merge the gene symbols with your top gene data
# Ensure the order matches the original ENSEMBL IDs
top_gene_symbols_e <- gene_mapping_e$hgnc_symbol[match(top_gene_ensembl_ids_e, gene_mapping_e$ensembl_gene_id)]

# Replace any missing symbols with the ENSEMBL IDs (in case symbols are missing)
top_gene_symbols_e[is.na(top_gene_symbols_e)] <- top_gene_ensembl_ids_e[is.na(top_gene_symbols_e)]

# Optional: Log transform the counts to reduce the dynamic range
log_top_counts_e <- log2(top_gene_counts_e + 1)

# Step 4: Ensure the annotation_col is a data frame with proper column names (for Control vs. EtOH)
annotation_col_e <- data.frame(Condition = colData(dds_filtered)$condition)
rownames(annotation_col_e) <- colnames(dds_filtered)

# Step 5: Save the heatmap as a PDF with gene symbols as row names for EtOH
pdf("r_heatmap_etoh_top_genes.pdf", width = 10, height = 8)

# Generate the heatmap with gene symbols as row names
pheatmap(log_top_counts_e, 
         cluster_rows = TRUE,     
         cluster_cols = TRUE,     
         show_rownames = TRUE,    # Show row names for the top genes (gene symbols)
         labels_row = top_gene_symbols_e,  # Use gene symbols as row labels
         show_colnames = TRUE,    # Keep column names
         annotation_col = annotation_col_e,   # Correctly formatted annotation data
         main = "R_Heatmap of Top 20 Significant Genes EtOH")

dev.off()  # Close the PDF device

```

## Top Genes Heatmap Control_NASH

```{r, warning=FALSE, message=FALSE}

# Assuming you already have `sigs_control_nash` with gene symbols loaded
# Load the DESeq2 object that contains the count data (dds_filtered object)
# Filter the DESeq2 object to only keep Control and NASH samples
dds_filtered_nash <- dds_filtered[, colData(dds_filtered)$condition %in% c("Control", "NASH")]

# Get the normalized counts for the significant genes from Control vs. NASH
# Extract only the genes that are significant and present in your filtered data
significant_genes <- rownames(sigs_nash)  # Assuming rownames are gene symbols or ENSEMBL IDs
normalized_counts <- counts(dds_filtered_nash, normalized = TRUE)
sig_gene_counts <- normalized_counts[significant_genes, ]


# Step 1: Sort the significant genes from NASH by log2FoldChange and select the top 20

# Select top 10 upregulated genes (highest log2FoldChange)
top_up_n <- sigs_nash_df[order(-sigs_nash_df$log2FoldChange), ][1:10, ]

# Select top 10 downregulated genes (lowest log2FoldChange)
top_down_n <- sigs_nash_df[order(sigs_nash_df$log2FoldChange), ][1:10, ]

# Combine top up- and downregulated genes
top_genes_nash <- rbind(top_up_n, top_down_n) 


# Step 2: Get the normalized counts for these top genes
top_gene_ensembl_ids_nash <- rownames(top_genes_nash)  # These are ENSEMBL IDs
top_gene_counts_nash <- normalized_counts[top_gene_ensembl_ids_nash, ]

# Step 3: Use biomaRt to convert ENSEMBL IDs to gene symbols for NASH


# Query to get gene symbols for the given ENSEMBL IDs
gene_mapping_nash <- getBM(
    filters = "ensembl_gene_id",
    attributes = c("ensembl_gene_id", "hgnc_symbol"),
    values = top_gene_ensembl_ids_nash,
    mart = mart
)

# Merge the gene symbols with your top gene data
# Ensure the order matches the original ENSEMBL IDs
top_gene_symbols_nash <- gene_mapping_nash$hgnc_symbol[match(top_gene_ensembl_ids_nash, gene_mapping_nash$ensembl_gene_id)]

# Replace any missing symbols with the ENSEMBL IDs (in case symbols are missing)
top_gene_symbols_nash[is.na(top_gene_symbols_nash)] <- top_gene_ensembl_ids_nash[is.na(top_gene_symbols_nash)]

# Optional: Log transform the counts to reduce the dynamic range
log_top_counts_nash <- log2(top_gene_counts_nash + 1)

# Step 4: Ensure the annotation_col is a data frame with proper column names (for Control vs. NASH)
annotation_col_nash <- data.frame(Condition = colData(dds_filtered)$condition)
rownames(annotation_col_nash) <- colnames(dds_filtered)

# Step 5: Save the heatmap as a PDF with gene symbols as row names for NASH
pdf("r_heatmap_nash_top_genes.pdf", width = 10, height = 8)

# Generate the heatmap with gene symbols as row names
pheatmap(log_top_counts_nash, 
         cluster_rows = TRUE,     
         cluster_cols = TRUE,     
         show_rownames = TRUE,    # Show row names for the top genes (gene symbols)
         labels_row = top_gene_symbols_nash,  # Use gene symbols as row labels
         show_colnames = TRUE,    # Keep column names
         annotation_col = annotation_col_nash,   # Correctly formatted annotation data
         main = "R_Heatmap of Top 20 Significant Genes NASH")

dev.off()  # Close the PDF device

```




## GSEA for Control_EtOH

```{r, warning=FALSE, message=FALSE}
# Create a ranked gene list based on log2FoldChange
# Remove any NA values and sort by log2FoldChange
gene_list <- sigs_etoh_annotated$log2FoldChange
names(gene_list) <- sigs_etoh_annotated$ensembl_gene_id  # Use gene symbols or ENSEMBL IDs

# Remove NA values and sort the list in decreasing order (for GSEA)
gene_list <- sort(na.omit(gene_list), decreasing = TRUE)

# Load necessary packages
library(clusterProfiler)
library(org.Hs.eg.db)  # Load human gene annotation

# Perform GSEA using GO terms (Biological Process, Molecular Function, and Cellular Component)
gsea_results <- gseGO(geneList = gene_list, 
                        OrgDb = org.Hs.eg.db,    # Use human gene annotation
                        ont = "BP",             # Perform GSEA for all GO categories (BP, MF, CC)
                        pvalueCutoff = 0.05,     # p-value cutoff for enrichment
                        keyType = "ENSEMBL",     # Use ENSEMBL IDs for genes
                        minGSSize = 15,          # Minimum size of gene sets
                        maxGSSize = 500,         # Maximum size of gene sets
                        eps = 0)


# Convert GSEA results to a data frame
gsea_df <- as.data.frame(gsea_results)

# Add a Regulation column indicating up- or down-regulation
gsea_df$Regulation <- ifelse(gsea_df$NES > 0, "Upregulated", "Downregulated")


# Modify the dot plot to color by Regulation (Upregulated or Downregulated)
go_gsea_plot <- ggplot(gsea_df[1:20, ], aes(x = NES, y = reorder(Description, NES), color = Regulation, size = setSize)) +
  geom_point() +
  scale_color_manual(values = c("Upregulated" = "red", "Downregulated" = "blue")) +
  labs(title = "Top 20 Enriched GO Terms R_ Control EtOH",
       x = "Normalized Enrichment Score (NES)",
       y = "GO Terms") +
  theme_minimal()

# Print the plot
print(go_gsea_plot)

# Save the plot as a PDF file
ggsave("r_go_gsea_plot_etoh.pdf", plot = go_gsea_plot, width = 10, height = 7)
```



### GSEA for Control_NASH

``` {r,warning=FALSE, message=FALSE}
# Create a ranked gene list based on log2FoldChange
# Remove any NA values and sort by log2FoldChange
gene_list <- sigs_nash_annotated$log2FoldChange
names(gene_list) <- sigs_nash_annotated$ensembl_gene_id  # Use gene symbols or ENSEMBL IDs

# Remove NA values and sort the list in decreasing order (for GSEA)
gene_list <- sort(na.omit(gene_list), decreasing = TRUE)

# Load necessary packages
library(clusterProfiler)
library(org.Hs.eg.db)  # Load human gene annotation

# Perform GSEA using GO terms (Biological Process, Molecular Function, and Cellular Component)
gsea_results <- gseGO(geneList = gene_list, 
                        OrgDb = org.Hs.eg.db,    # Use human gene annotation
                        ont = "BP",             # Perform GSEA for all GO categories (BP, MF, CC)
                        pvalueCutoff = 0.05,     # p-value cutoff for enrichment
                        keyType = "ENSEMBL",     # Use ENSEMBL IDs for genes
                        minGSSize = 15,          # Minimum size of gene sets
                        maxGSSize = 500,         # Maximum size of gene sets
                        eps = 0)


# Convert GSEA results to a data frame
gsea_df <- as.data.frame(gsea_results)

# Add a Regulation column indicating up- or down-regulation
gsea_df$Regulation <- ifelse(gsea_df$NES > 0, "Upregulated", "Downregulated")


# Modify the dot plot to color by Regulation (Upregulated or Downregulated)
go_gsea_plot_nash <- ggplot(gsea_df[1:20, ], aes(x = NES, y = reorder(Description, NES), color = Regulation, size = setSize)) +
  geom_point() +
  scale_color_manual(values = c("Upregulated" = "red", "Downregulated" = "blue")) +
  labs(title = "Top 20 Enriched GO Terms R_Control NASH",
       x = "Normalized Enrichment Score (NES)",
       y = "GO Terms") +
  theme_minimal()

# Print the plot
print(go_gsea_plot_nash)

# Save the plot as a PDF file
ggsave("r_go_gsea_plot_control_nash.pdf", plot = go_gsea_plot_nash, width = 10, height = 7)

```



